Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism

### 2024-08-06 _tex/index.tex
Built site for gh-pages

Nevertheless, the structural approach offers an additional key advantage
over the potential outcomes approach: it enables the graphical
representation of any system through directed acyclic graphs
\citep[DAG,][]{Gross_et_al_2018, Neal_2020}. DAGs serve as heuristics,
effectively conveying the assumed causal structure of a system. A
heuristic is a strategy that simplifies information to make decisions
more quickly, efficiently, and sometimes more accurately than complex
methods \citep{Chow_2015}. Consequently, DAGs do not represent detailed
statistical models but allow researchers to deduce which statistical
models can provide valid causal inferences, assuming the causal
structure depicted in the DAG is accurate \citep{McElreath_2020}.

@@

DAGs offer two key advantages for modeling causality. Firstly, they
represent causal relations in a nonparametric and fully interactive
manner. This generality implies that feasible causal analysis strategies
usually do not require specifying the type of data or the nature of the
functional dependence among variables \citep{Morgan_et_al_2014}.
Secondly, regardless of complexity, DAGs can represent various causal
structures using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. These building blocks can be depicted
in three ways: the magnified representation, the standard
representation, and the structural causal model form
\citep[SCM,][]{Morgan_et_al_2014}.

@@

Lastly, the right panels of Figure~\ref{fig-dags_scms} depict the SCM
representation of the five fundamental building blocks. SCMs are formal
mathematical models that represent causal relationships within a system
or population \citep{Hitchcock_2024}. An SCM is defined by the set of
endogenous variables \(V\), the set of exogenous variables \(E\), and
the set of functions \(F=\{f_{X},f_{Z},f_{Y}\}\)
\citep{Pearl_2009, Neal_2020}. These functions, referred to as
structural equations, specify each endogenous variable as a
nonparametric function of other variables. Moreover, SCMs use the symbol
`\(:=\)' to indicate the variables' asymmetrical causal dependence and
the symbol `\(\perp\!\!\!\perp\)' to represent \emph{d-separation},
which roughly equates to the concept of variable independence. The
concepts of d-separation and causal (in)dependence are explored in the
following section.

@@

A careful examination of Figure~\ref{fig-dags_scms} highlights the
assumptions underlying these building blocks. Figures
\ref{fig-mdag_bb1}, \ref{fig-sdag_bb1}, and SCM \ref{fig-scm_bb1} depict
two unconnected nodes, representing a scenario where variables \(X\) and
\(Y\) are not causally related. Figures \ref{fig-mdag_bb2},
\ref{fig-sdag_bb2}, and SCM \ref{fig-scm_bb2} illustrate two connected
nodes, showing a scenario where a \emph{parent} node \(X\) exerts a
causal influence on a \emph{child} node \(Y\). Consequently, \(Y\) is
considered a \emph{descendant} of \(X\). Figures \ref{fig-mdag_bb3},
\ref{fig-sdag_bb3}, and SCM \ref{fig-scm_bb3} depict a \emph{chain} or
\emph{pipe}, where \(X\) influences \(Z\), and \(Z\) influences \(Y\).
In this configuration, \(X\) is a parent node of \(Z\), and \(Z\) is a
parent node of \(Y\). Furthermore, the DAGs show that \(X\) is an
\emph{ancestor} of \(Y\) and that \(Z\) fully \emph{mediates} the
relationship between these variables. Figures \ref{fig-mdag_bb4},
\ref{fig-sdag_bb4}, and SCM \ref{fig-scm_bb4} illustrate a \emph{fork},
where variables \(X\) and \(Y\) are both influenced by \(Z\). Here,
\(Z\) is a parent node of \(X\) and \(Y\). Finally, Figures
\ref{fig-mdag_bb5}, \ref{fig-sdag_bb5}, SCM \ref{fig-scm_bb5} depict a
\emph{collider}, also known as \emph{inmorality}, where variables \(X\)
and \(Y\) are concurrent causes of \(Z\). In this configuration, \(X\)
and \(Y\) are not causally related to each other but both influence
\(Z\). Additionally, in all SCMs, the errors are assumed to be mutually
independent of each other and from all other variables in the graph, as
evidenced by the pairwise relations \(e_{X} \perp\!\!\!\perp e_{Y}\),
\(e_{X} \perp\!\!\!\perp e_{Z}\), and \(e_{Z} \perp\!\!\!\perp e_{Y}\).

Given the heuristic nature of DAGs, a motivating example can clarify how
to use the five fundamental building blocks to construct the causal
structure of a system. Consider a research problem where the causal
effect of a variable \(T\) on an outcome \(Y\) needs investigation. The
problem suggests that a variable \(X\) potentially influences both \(T\)
and \(Y\). Beyond these relationships, the problem does not specify any
further variables of interest. Such scenarios are not hard to imagine.
For instance, \(T\) might represent different treatments that could
affect the recovery from cancer \(Y\), while \(X\) could denote the
patient's age. Similarly, in the context of a DCJ study like the one
described by \citet{Boonen_et_al_2020}, \(T\) could represent the
duration of a child's cochlear implant use, which might influence the
child's overall speech quality \(Y\), with \(X\) indicating the child's
hearing status. {(not a bad example, but I prefer one using writing
skills)}

Figure~\ref{fig-example} illustrates the plausible causal structure of
the motivating example. A detailed examination of Figures
\ref{fig-mdag_example1}, \ref{fig-sdag_example1}, and SCM
\ref{fig-scm_example1} reveals the presence of at least four of the five
fundamental building blocks. The figures display multiple descendants,
as indicated by pairwise relations such as \(X \rightarrow T\),
\(X \rightarrow Y\), and \(T \rightarrow Y\). Additionally, the figures
features multiple pairs of unconnected nodes, evident from the relations
\(e_{T} \perp\!\!\!\perp e_{X}\), \(e_{T} \perp\!\!\!\perp e_{Y}\), and
\(e_{X} \perp\!\!\!\perp e_{Y}\). Finally, the figures illustrate a fork
with \(X \rightarrow \{T,Y\}\), as well as colliders such as
\(\{X,e_{T}\} \rightarrow T\) and \(\{X,T,e_{Y}\} \rightarrow Y\).

>>>>

However, the structural approach offers an additional key advantage over
the potential outcomes approach: it enables the graphical representation
of any system through directed acyclic graphs
\citep[DAG,][]{Gross_et_al_2018, Neal_2020}. DAGs serve as heuristics,
effectively conveying the assumed causal structure of a system. A
heuristic is a strategy that simplifies information to make decisions
more quickly, efficiently, and sometimes more accurately than complex
methods \citep{Chow_2015}. Consequently, DAGs do not represent detailed
statistical models but allow researchers to deduce which statistical
models can provide valid causal inferences, assuming the causal
structure depicted in the DAGs are accurate \citep{McElreath_2020}.

@@

DAGs offer two key advantages for modeling causality. Firstly, DAGs
represent causal relations in a nonparametric and fully interactive
manner. This feature implies that feasible causal analysis strategies
usually do not require specifying the type of data or the nature of the
functional dependence among variables \citep{Morgan_et_al_2014}.
Secondly, regardless of complexity, DAGs can represent various causal
structures using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. These building blocks can be depicted
in three ways: the magnified representation, the standard
representation, and the structural causal model form
\citep[SCM,][]{Morgan_et_al_2014}.

@@

Lastly, the right panels of Figure~\ref{fig-dags_scms} depict the SCM
form of the five fundamental building blocks. SCMs are formal
mathematical models that represent causal relationships within a system
or population \citep{Hitchcock_2024}. An SCM is defined by a set of
endogenous variables \(V\), a set of exogenous variables \(E\), and a
set of functions \(F=\{f_{X},f_{Z},f_{Y}\}\)
\citep{Pearl_2009, Neal_2020}. These functions, referred to as
structural equations, specify each endogenous variable as nonparametric
functions of other variables. Moreover, SCMs use the symbol `\(:=\)' to
indicate the variables' asymmetrical causal dependence and the symbol
`\(\perp\!\!\!\perp\)' to represent \emph{d-separation}, which roughly
equates to the concept of variable independence. The concepts of
d-separation and causal (in)dependence are explored in the following
section.

@@

A careful examination of Figure~\ref{fig-dags_scms} highlights the
assumptions underlying these building blocks. Figures
\ref{fig-mdag_bb1}, \ref{fig-sdag_bb1}, and SCM \ref{fig-scm_bb1} depict
two unconnected nodes, representing a scenario where variables \(X\) and
\(Y\) are not causally related. Figures \ref{fig-mdag_bb2},
\ref{fig-sdag_bb2}, and SCM \ref{fig-scm_bb2} illustrate two connected
nodes, showing a scenario where a \emph{parent} node \(X\) exerts a
causal influence on a \emph{child} node \(Y\). Consequently, \(Y\) is
considered a \emph{descendant} of \(X\). Figures \ref{fig-mdag_bb3},
\ref{fig-sdag_bb3}, and SCM \ref{fig-scm_bb3} depict a \emph{chain} or
\emph{pipe}, where \(X\) influences \(Z\), and \(Z\) influences \(Y\).
In this configuration, \(X\) is a parent node of \(Z\), and \(Z\) is a
parent node of \(Y\). Furthermore, the DAGs show that \(X\) is an
\emph{ancestor} of \(Y\) and that \(Z\) fully \emph{mediates} the
relationship between these variables. Figures \ref{fig-mdag_bb4},
\ref{fig-sdag_bb4}, and SCM \ref{fig-scm_bb4} illustrate a \emph{fork},
where variables \(X\) and \(Y\) are both influenced by \(Z\). Here,
\(Z\) is a parent node of \(X\) and \(Y\). Finally, Figures
\ref{fig-mdag_bb5}, \ref{fig-sdag_bb5}, SCM \ref{fig-scm_bb5} depict a
\emph{collider}, also known as \emph{inmorality}, where variables \(X\)
and \(Y\) are concurrent causes of \(Z\). In this configuration, \(X\)
and \(Y\) are not causally related to each other but both influence
\(Z\). Additionally, in all SCMs, the errors are assumed to be mutually
independent of each other and of all other variables in the graph, as
evidenced by the pairwise relations \(e_{X} \perp\!\!\!\perp e_{Y}\),
\(e_{X} \perp\!\!\!\perp e_{Z}\), and \(e_{Z} \perp\!\!\!\perp e_{Y}\).

Given the heuristic nature of DAGs, a motivating example can help
clarify how to use the five fundamental building blocks to construct the
causal structure of a system. Consider a research problem where the
causal effect of a variable \(T\) on an outcome \(Y\) needs
investigation. The problem suggests that a variable \(X\) potentially
influences both \(T\) and \(Y\). Beyond these relationships, the problem
does not specify any further variables of interest. Such scenarios are
not hard to imagine. For instance, \(T\) might represent different
treatments that could affect the recovery from cancer \(Y\), while \(X\)
could denote the patient's age. Similarly, in the context of a DCJ study
like the one described by \citet{Boonen_et_al_2020}, \(T\) could
represent the duration of a child's cochlear implant use, which might
influence the child's overall speech quality \(Y\), with \(X\)
indicating the child's hearing status. {(not a bad example, but I prefer
one using writing skills)}

Figure~\ref{fig-example} illustrates the plausible causal structure of
the motivating example. A detailed examination of Figures
\ref{fig-mdag_example1}, \ref{fig-sdag_example1}, and SCM
\ref{fig-scm_example1} reveals the presence of at least four of the five
fundamental building blocks. The figures display multiple descendants,
as indicated by pairwise relations such as \(X \rightarrow T\),
\(X \rightarrow Y\), and \(T \rightarrow Y\). Additionally, the figures
features multiple pairs of unconnected nodes, evident from the relations
\(e_{T} \perp\!\!\!\perp e_{X}\), \(e_{T} \perp\!\!\!\perp e_{Y}\), and
\(e_{X} \perp\!\!\!\perp e_{Y}\). Finally, the figures illustrate the
fork \(X \rightarrow \{T,Y\}\), and two colliders with
\(\{X,e_{T}\} \rightarrow T\) and \(\{X,T,e_{Y}\} \rightarrow Y\).
